---
title: "Personal Movement Monitoring"
output: html_document
---

It is now possible to collect a large amount of data about personal movement
using activity monitoring devices such as a Fitbit, Nike Fuelband, or Jawbone
Up. These type of devices are part of the "quantified self" movement - a group
of enthusiasts who take measurements about themselves regularly to improve their
health, to find patterns in their behavior, or because they are tech geeks. But
these data remain under-utilized both because the raw data are hard to obtain
and there is a lack of statistical methods and software for processing and
interpreting the data.

This assignment makes use of data from a personal activity monitoring device.
This device collects data at 5 minute intervals through out the day. The data
consists of two months of data from an anonymous individual collected during the
months of October and November, 2012 and include the number of steps taken in 5
minute intervals each day.

We are going to answer some questions about the personal movement monitoring
data (which we stored in our /data directory):
```{r}   
setwd("C:/Users/Fer/Desktop/Coursera")
activity <- read.csv("./data/activity.csv")
activity$date <- as.Date (activity$date)
```

Note dates input will be taken as.Date in order to perform a better analysis.

What is mean total number of steps taken per day?
--------------------------------------------------------------
Lets show that by making a histogram of the total number of steps taken each day


```{r, warning=FALSE}
library(ggplot2)
dailysteps <- aggregate(steps ~ date, activity, sum, na.rm=TRUE)
hist1 <- ggplot(dailysteps,aes(x=date,y=steps)) +
        geom_bar(stat="identity") +
        ggtitle("steps per day")
        
print(hist1)
```

Now we can calculate the mean and median for the steps taken per day

```{r}
c(mean(dailysteps$steps), median(dailysteps$steps))
```

What is the average daily activity pattern?
--------------------------------------------------------------
The same strategy can be performed for interval puposes:
```{r}   
stepsPerinterval <- aggregate(steps ~ interval, activity, sum)       
```
Let's show a time series plot of the 5-minute intervals and the average steps
taken
```{r}
plot(stepsPerinterval, type = "l", col = "blue")
```


Now we can see the maximum step containing interval is the next one:

```{r}
stepsPerinterval[which.max (stepsPerinterval$steps),]
```


Imputing missing values
--------------------------------------------------------------
We can take a look at how many missing values are there:
```{r}
sum (is.na(activity$steps))
```
And we are going to avoid them by replacing them with the the mean for their
5-minute interval and including them in a new dataset:
```{r}       
meanstepsinterval <- aggregate(steps ~ interval, activity, mean) 
nonullactivity <- activity  
for (i in 1:nrow(activity)){
    if(is.na(activity$steps[i])){
        nonullactivity$steps[i]<- meanstepsinterval[meanstepsinterval$interval==activity$interval[i],]$steps
    }
}
```

Now, let's show the result on a histogram:

```{r}
dailysteps1 <- aggregate(steps ~ date, nonullactivity, sum, na.rm=TRUE)
hist2 <- ggplot (dailysteps1, aes(x=date, y=steps)) + geom_bar(stat="identity") + ggtitle("Total number of steps per day")
print (hist2)
```

We can see as expected the mean does not differ, but the median is alittle lower:

```{r}
c(mean(dailysteps1$steps), median(dailysteps1$steps))
```


Are there differences in activity patterns between weekdays and weekends?
--------------------------------------------------------------

Let's show it: First we can adapt the dataset to include weekend indicator:
```{r}  
nonullactivity$weekend <- sapply (nonullactivity$date, FUN= function (x) {return (weekdays (x)=="domingo" | weekdays (x)=="sábado")})
nonullactivity$weekend<- ifelse (nonullactivity$weekend, "weekend", "weekday")
```

And here is the plot:

```{r}
activityintdaytype <- aggregate (steps~interval + weekend, nonullactivity, mean)
ggplot (activityintdaytype, aes(x=interval, y=steps)) + geom_bar(stat="identity") + facet_wrap (~weekend, nrow=2, ncol=1) + theme_bw() + geom_line (color="red", size=1)
```

We can see there are differences on the average steps per interval in weekdays and weekends.
